Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-05-08 paper/paper.Rmd
fix all references and upd paper.bib

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich \AND'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].

@@

This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)). 
The earliest of these implementations is the \pkg{sperrorest} package [@sperrorest], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks. 
Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].

@@

In cross-validation for predictive model assessment, we consider the following formal setting. 
We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)}\in\mathbb{R}$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$. Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
\[
\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population. 
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1996]. 
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures. 
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand1996 for classification, and ref?? for regression tasks].

@@

A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie2015], sometimes with a focus on the spatial distribution of LOO error [@willmott2006]. 
Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006]. 
Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005]. 
One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005]. 
When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]. 
Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].

@@

With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modelling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr} with suitable algorithms. 
In this context, \pkg{mlr3} stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; mlr3book]. 
All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.

@@

The "Spatial CV" method after [@sperrorest] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
However, it does not make use of arbitrary features but uses the coordinates of all observations to create clusters in the spatial domain.
This is what lead to the naming of `spcv_coords` in \pkg{mlr3spatiotempcv}.

@@

Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operate in a two-dimensional space.

@@

<!-- spcv coords: mention generic name -->
In @sperrorest, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.
Hence, in \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich \AND'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., [@cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005, @brenning2012; @pohjankukka2017; @roberts2017].

@@

This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)). 
The earliest of these implementations is the \pkg{sperrorest} package [@brenning2012], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks. 
Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].

@@

In cross-validation for predictive model assessment, we consider the following formal setting. 
We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)}\in\mathbb{R}$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$. Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
\[
\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population. 
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000]. 
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures. 
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand1996 for classification, and ref?? for regression tasks].

@@

A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006]. 
Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006]. 
Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005]. 
One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005]. 
When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]. 
Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].

@@

With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms. 
In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book]. 
All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.

@@

The "Spatial CV" method after [@brenning2012] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
However, it does not make use of arbitrary features but uses the coordinates of all observations to create clusters in the spatial domain.
This is what lead to the naming of `spcv_coords` in \pkg{mlr3spatiotempcv}.

@@

Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operate in a two-dimensional space.

@@

<!-- spcv coords: mention generic name -->
In @brenning2012, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.
Hence, in \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
